---
title: "Assignment-1"
output:
  html_document: default
  pdf_document: default
  word_document: default
---



```{r}


library(ggplot2)
library(plyr)

activity<-read.csv("activity.csv")

#### Processing the Data

activity$day<-weekdays(as.Date(activity$date))
activity$DateTime<-as.POSIXct(activity$date,format="%Y-%m-%d")
clean<-activity[!is.na(activity$steps),]
```
What is mean total number of steps taken per day ?

Calculate the total number of steps taken per day

```{r}

### calculate total steps per date 

sumTable<-aggregate(activity$steps~activity$date,FUN = sum,)

colnames(sumTable)<-c("Date", "steps")
```

Make a histogram of the total number of steps taken each day

```{r}

hist(sumTable$steps,breaks = 5,xlab="steps",main="Total steps per day")


```

Calculate and report the mean and median of the total number of steps taken per day 

```{r}

## Mean of steps

as.integer(mean(sumTable$steps))

```


```{r}
as.integer(median(sumTable$steps))

```

The average number of steps taken each day was 10766 steps
The median number of steps taken each day was 10765 steps

Make a time series plot(i.e type="l") of the 5-minimum interval (x-axis) and the average number of steps taken, averaged across all days(y-axis)


```{r}
library(plyr)
library(ggplot2)

## pulling data without nas 

clean<-activity[!is.na(activity$steps),]

## create average number of steps per interval

intervalTable<-ddply(clean,.(interval),summarize,Avg=mean(steps))

## create line plot of average number of steps per interval 

p<-ggplot(intervalTable,aes(x=interval,y=Avg),xlab="Interval",ylab="Average Number of Steps")
p+geom_line()+xlab("Interval")+ylab("Average Number of steps")+ggtitle("Average Number of Steps per Interval")




```

Which 5-minute interval on average across all the days in the dataset,contains the maximum number of steps ?

```{r}
## Maximum step by interval

maxSteps<-max(intervalTable$Avg)

## Which Interval contain maximum number average number of steps

intervalTable[intervalTable$Avg==maxSteps,1]


```
 
The maximum number of steps for a  5-minute interval was 206 steps.
the 5-minute interval which had the maximum number of steps was the 835 interval.

Calculate and report the total number of missing values in the dataset (i.e the total number of rows with NAs)

```{r}

## Inputing missing values 

nrow(activity[is.na(activity$steps),])



```

The total number of rows with steps='NA' is 2304

Device a strategy for filling in all of the missing values in the dataset. The strategy doesnot need to be sophisticated. For example ,you could use the mean/median for that day,or the mean for that 5- minute interval ,etc

My strategy for filling in NAs will be to subsitute the missing steps  with the average 5-minute  interval based on the day of the week 

```{r}

## Create  the average number of steps per weekday and interval

avgTable<-ddply(clean,.(interval,day),summarize,Avg=mean(steps))

## create dataset with all NAs for substitution

nadata<-activity[is.na(activity$steps),]


## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))

newdata2 <- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

##Merge the NA averages and non NA data together
mergeData <- rbind(clean, newdata2)

```
Create a new dataset that is equal to the orignal dataset but with the missing data filled in.

```{r}
##Create sum of steps per date to compare with step 1
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")

## Mean of Steps with NA data taken care of
as.integer(mean(sumTable2$Steps))

```

```{r}
## Median of Steps with NA data taken care of
as.integer(median(sumTable2$Steps))

```




```{r}

## Creating the histogram of total steps per day, categorized by data set to show impact

hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")

hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)

legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )


```

The new mean of the imputed data is 10821 steps compared to the old mean of 10766 steps. That creates a difference of 55 steps on average per day.

The new median of the imputed data is 11015 steps compared to the old median of 10765 steps. That creates a difference of 250 steps for the median.

However, the overall shape of the distribution has not changed.


```{r}

## Create new category based on the days of the week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r}

library(lattice)

## Warning: package 'lattice' was built under R version 3.1.3
## Summarize data by interval and type of day
intervalTable2<-ddply(mergeData,.(interval,day),summarize, Avg=mean(steps))

##Plot data in a panel plot
xyplot(Avg~interval|day, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```


